Selecting and ranking interesting pairs of cyclic temporal granularities

### 2020-11-19 main.tex
normalisation result update

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here
\setlength {\marginparwidth }{2cm}

\usepackage{mathtools,amssymb,booktabs,longtable,todonotes,amsthm} \def\mod{~\text{mod}~}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf Visualizing probability distributions across bivariate cyclic temporal granularities}

  \author{
        Sayani Gupta \thanks{Email: \href{mailto:Sayani.Gupta@monash.edu}{\nolinkurl{Sayani.Gupta@monash.edu}}} \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Rob J Hyndman \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Dianne Cook \\
    Department of Econometrics and Business Statistics, Monash University\\
     and \\     Antony Unwin \\
    University of Augsburg\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Visualizing probability distributions across bivariate cyclic temporal granularities}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
Deconstructing a time index into time granularities can assist in exploration and automated analysis of large temporal data sets. This paper describes several classes of time deconstructions using linear and cyclic time granularities. Linear time granularities respect the linear progression of time such as hours, days, weeks and months with respect to a baseline. Cyclic time granularities can be circular such as hour of the day, quasi-circular such as day of the month, and aperiodic such as public holidays. The hierarchical structure of granularities creates a nested ordering. Hour of the day and second of the minute are single-order-up. Hour of the week is multiple-order-up, because it passes over day of the week. Methods are provided for creating all possible granularities for a time index. A recommendation algorithm provides an indication whether a pair of granularities can be meaningfully examined together, called a harmony or when they cannot, called a clash.

The time granularities can be used to create visualizations of the data to explore for periodicities, associations and anomalies. The granularities can be considered to be categorical variables (ordered or unordered) which induces a grouping of the observations. Assuming a numeric response variable, the resulting graphics are then displays of distributions compared across combinations of categorical variables. A recommendation of appropriate distribution display is provided.

The methods are implemented in the open source R package \texttt{gravitas}. The functions for creating granularities and exploring the associated time series are consistent with a tidy workflow (\citet{Grolemund2018-po}), and the probability distributions can be examined using the range of graphics available in \texttt{ggplot2} \citep{Wickham2009pk}.
\end{abstract}

\noindent%
{\it Keywords:} data visualization, statistical distributions, time granularities, calendar algebra, periodicties, grammar of graphics, R
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Temporal data are available at various resolutions depending on the context. Social and economic data like GDP is often collected and reported at coarse temporal scales such as monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded at much finer temporal scales. Energy consumption may be collected every half an hour, energy supply may be collected every minute, and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales such as days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities \citep{aigner2011visualization}.

It is important to be able to navigate through all of these time granularities to have multiple perspectives on the periodicity of the observed data. This idea aligns with the notion of EDA \citep{Tukey1977-jx} which emphasizes the use of multiple perspectives on data to help formulate hypotheses before proceeding to hypothesis testing. Visualizing probability distributions conditional on one or more granularities is an indispensable tool for exploration. Analysts are expected to comprehensively explore the many ways to view and consider temporal data. However, the plethora of choices and the lack of a systematic approach to do so quickly becomes overwhelming.

Calendar-based graphics \citep{wang2020calendar} are useful in visualizing patterns in the weekly and monthly structure, and are helpful when checking for the effects of weekends or special days. Any temporal data at sub-daily resolution can also be displayed using this type of faceting \citep{Wickham2009pk} with days of the week, month of the year, or another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions. There can be many different time deconstructions, based on the calendar or on categorizations of time granularities.

Linear time granularities respect the linear progression of time and are non-repeating such as hours, days, weeks and months. One of the first attempts to characterize these granularities is due to \citet{Bettini1998-ed}. However, the definitions and rules defined are inadequate for describing cyclic or repeating granularities. Hence, there is a need to define some new cyclic time granularities, that can be useful in visualizations. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular granularities are hour of the day and day of the week; an example of a quasi-circular granularity is day of the month; examples of aperiodic granularities are public holidays and school holidays.

Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities such as second of the minute, or multiple-order-up granularities such as second of the hour. The lubridate package \citep{Grolemund2011-vm} provides tools to access and manipulate common date-time objects. However, most of its accessor functions are limited to single-order-up granularities.

The motivation for this work stems from the desire to provide methods to better understand large quantities of measurements on energy usage reported by smart meters in households across Australia, and indeed many parts of the world. Smart meters currently provide half-hourly use in kWh for each household, from the time that they were installed, some as early as 2012. Households are distributed geographically and have different demographic properties such as the existence of solar panels, central heating or air conditioning. The behavioral patterns in households vary substantially, for example, some families use a dryer for their clothes while others hang them on a line, and some households might consist of night owls, while others are morning larks. It is common to see aggregates \citep[see][]{Goodwin_2012} of usage across households, such as half-hourly total usage by state, because energy companies need to plan for maximum loads on the network. But studying overall energy use hides the distribution of usage at finer scales, and makes it more difficult to find solutions to improve energy efficiency. We propose that the analysis of smart meter data will benefit from systematically exploring energy consumption by visualizing the probability distributions across different deconstructions of time to find regular patterns/anomalies. Although, the motivation came through the smart meter example, this is a problem that is relevant to any temporal data observed more than once per year.

This work provides tools for systematically exploring bivariate granularities within the tidy workflow. In particular, we

\begin{itemize}
\tightlist
\item
  provide a formal characterization of cyclic granularities;
\item
  facilitate manipulation of single- and multiple-order-up time granularities through cyclic calendar algebra;
\item
  develop an approach to check the feasibility of creating plots or drawing inferences for any two cyclic granularities;
\item
  recommend prospective probability distributions for exploring distributions of a univariate dependent variable across pair of granularities.
\end{itemize}

The remainder of the paper is organized as follows: Section \ref{sec:linear-time} provides some background material on linear granularities and calendar algebra for computing different linear granularities. Section \ref{sec:cyclic-gran} formally characterizes different cyclic time granularities by extending the framework of linear time granularities. Section \ref{sec:cyclic-calendar} introduces cyclic calendar algebra for computing cyclic time granularities. Section \ref{sec:data-structure} discusses the data structure for exploring the conditional distributions of the associated time series across pairs of cyclic time granularities. Section \ref{sec:visualization} discusses the role of different factors in constructing an informative and trustworthy visualization. Section \ref{sec:application} examines how systematic exploration can be carried out for a temporal and non-temporal application. Section \ref{sec:discussion} summarizes this paper and discusses possible future direction.

\hypertarget{sec:linear-time}{%
\section{Linear time granularities}\label{sec:linear-time}}

Discrete abstraction of time such as weeks, months or holidays can be thought of as ``time granularities''. Time granularities are \textbf{linear} if they respect the linear progression of time. There have been several attempts to provide a framework for formally characterizing time granularities, including \citet{Bettini1998-ed} which forms the basis of the work described here.

\hypertarget{definitions}{%
\subsection{Definitions}\label{definitions}}

\newtheorem{definition}{Definition}

\begin{definition}\label{def:definition}
A {\bf time domain} is a pair $(T; \le)$ where $T$ is a non-empty set of time instants and $\le$ is a total order on $T$.
\end{definition}

The time domain is assumed to be \emph{discrete}, and there is unique predecessor and successor for every element in the time domain except for the first and last.

\begin{definition}\label{def:index set}
There is a unique {\bf index set}, $Z=\{z: z \in \mathbb{Z}_{\geq 0}\}$, which map the time instants to a set of non-negative integers. Often, this is thought of as $t=0, \dots, T$.
\end{definition}

\begin{definition}\label{def:linear}
A {\bf linear granularity} is a mapping $G$ from the integers (the index set, $Z$) to subsets of the time domain such that:
  (1) if $i < j$ and $G(i)$ and $G(j)$ are non-empty, then each element of $G(i)$ is less
than all elements of $G(j)$; and
  (2) if $i < k < j$ and $G(i)$ and $G(j)$ are non-empty, then $G(k)$ is non-empty.
Each non-empty subset $G(i)$ is called a {\bf granule}.
\end{definition}

\noindent This implies that the granules in a linear granularity are non-overlapping, continuous and ordering is maintained. The indexing for each granule can also be associated with textual representation, called the label. A discrete time model often uses a fixed smallest linear granularity named by \citet{Bettini1998-ed} \textbf{bottom granularity}. \autoref{fig:linear-time} illustrates the linear time granularities. Here, ``hour'' is the bottom granularity and ``day'', ``week'', ``month'' and ``year'' are linear granularities which maps from index set to subsets of the hourly time domain. If we have ``hour'' running from \(\{0, 1, \dots,t\}\), we will have ``day'' running from \(\{0, 1, \dots, \lfloor t/24\rfloor\}\). These linear granularities are ordered with ordering guided by the index set which is a set of integers. Hence, they are uni-directional and non-repeating.

@@

{\centering \includegraphics[width=1\linewidth]{Figs/linear-ex}

@@

\caption{Illustration of time domain, linear granularities and index set. Hour, day, week, month and year are linear granularities and can also be considered to be time domains. These are ordered with ordering guided by integers and hence is unidirectional and non-repeating. Hours could also be considered the index set, and a bottom granularity.}\label{fig:linear-time}
\end{figure}

\hypertarget{relativities}{%
\subsection{Relativities}\label{relativities}}

Properties of pairs of granularities fall into various categories.

\begin{definition}\label{def:finerthan}
A linear granularity $G$ is {\bf finer than} a linear granularity $H$, denoted $G \preceq H$, if for each index $i$, there exists an index $j$ such that
$G(i) \subset H(j).$
\end{definition}

\begin{definition}\label{def:groupsinto}
A linear granularity $G$ {\bf groups into} a linear granularity $H$, denoted
$G \trianglelefteq H$, if for each index $j$ there exists a (possibly infinite) subset $S$ of the integers such that $H(j) = \bigcup_{i \in S}G(i).$
\end{definition}

\newtheorem*{example}{Example}

\begin{example}
{\rm Both $day \trianglelefteq week$ and $day \preceq week$ holds, since every granule of $week$ is the union of some set of granules of day and each day is a subset of a $week$. Consider another example where $day \trianglelefteq month$. This relationship however is incomplete without its association to periodicity. Each month is a grouping of the same number of days over years, hence the period of the grouping $(day, month)$ is one year, if leap years are ignored. This grouping period becomes 4 and 400 years with the inclusion of leap years and leap seconds respectively.}
\end{example}

\begin{definition}\label{def:periodical}
A granularity $G$ is {\bf periodical} with respect to a granularity $H$ if:
(1) $G \trianglelefteq H$; and
(2) there exist $R$, $P \in \mathbb{Z}_+$, where $R$ is less than the number of granules of $H$, such that for all $i \in \mathbb{Z}$, if $H(i) = \bigcup_{j \in S}G(j)$ and $H (i + R) \neq \phi$ then $H (i + R) = \bigcup_{j \in S} G(j + P)$.
\end{definition}

If \({S_0,\dots,S_{R-1}}\) are the sets of indexes of \(G\) describing
\(H(0), \dots, H(R - 1)\), respectively, then the description of an arbitrary granule \(H(j)\) is given by \(\bigcup_{i \in S_j ~\text{mod}~R}G(P*\lfloor j/R \rfloor + i)\). Also, granularities can be periodical with respect to other granularities, except for a finite number of spans of time where they behave in an anomalous way, called \textbf{quasi-periodic} relationships by \citet{Bettini2000-vy}.

\begin{example}
{\rm In a Gregorian calendar without leap years we could say $day$ groups periodically into month with the period $P = 365$ and the number of granules of $month$ in each period given by $R=12$. In a Gregorian calendar with leap years, $day$ groups quasi-periodically into $month$ with the exceptions of the time domain corresponding to $29^{\text{th}}$ February of any year.}
\end{example}

\begin{definition}\label{def:order}
The {\bf order} of a linear granularity is the level of coarseness associated with a linear granularity. A linear granularity G will have lower order than H if each granule of G is composed of lower number of granules of bottom granularity than each granule of H.
\end{definition}

\begin{example}
{\rm For two linear granularities $G$ and $H$, if $G$ {\em groups into} or {\em finer than} $H$ then $G$ is of lower order than $H$. Moreover, if we consider bottom granularity as day, linear granularity week will have lower order than month since each week consist of less number of days than each month. Here neither  week {\em groups into} month or  week {\em finer than} month, but their relative order could be determined.}
\end{example}

@@

Granules in bottom granularity or any finer granularity may be aggregated in some manner to form larger granules belonging to a coarser granularity. A system of multiple granularities in lattice structures is referred to as a \textbf{calendar} by \citet{Dyreson_2000}. Linear time granularities are computed through an algebraic representation for time granularities, which is referred to as calendar algebra \citep{Ning_2002}. It is assumed that there exists a \emph{bottom granularity} and calendar algebra operations are designed to generate new granularities recursively from the bottom one. Some relevant calendar algebra operations are discussed below; these will be used in Section \ref{sec:cyclic-gran} for illustrations in circular and quasi-circular granularities.

\begin{definition}\label{def:grouping_operation}
Let $G_1$ be a full-integer labelled granularity, and $m$ a positive integer. {\bf The grouping operation} Group$_m(G)$ generates a new granularity $G_2$, by partitioning the granules of $G_1$ into $m$-granule groups and making each group a granule of the resulting granularity. More precisely, $G_2 = \text{Group}_m(G_1)$ is the full-integer labelled granularity such that for each integer $i$,
$$G_2(i) = \bigcup\limits_{j = (i-1)m+1}^{im} G_1(j).$$
The grouping operation Group$_m(G)$, provides a new granularity whose granules are the integer part of the quotient when the integer set spanned by linear granularity $G$ is divided by $m$, such that, $G_2(i) = \lfloor G_1(j)/m \rfloor$.
\end{definition}

\begin{example} 
{\rm Due to even length of $day$ and $week$, we can derive them from $hour$ using the grouping operation as follows: day = G$_{24}(\text{hour})$, week = G$_{24*7}(\text{hour})$.}
\end{example}

\noindent For more variations of calendar algebra operations, see \citet{Ning_2002}.

\hypertarget{sec:cyclic-gran}{%
\section{Cyclic time granularities}\label{sec:cyclic-gran}}

Cyclic granularities represent cyclical repetitions in time. They can be thought of as additional categorizations of time that are not linear. Cyclic granularities can be constructed by operating on two linear granularities. Cycles can be either \emph{regular}, called \textbf{circular}, or \emph{irregular}, \textbf{quasi-circular} when these two linear granularities relate periodically.

\hypertarget{sec:circular-gran-def}{%
\subsection{Circular}\label{sec:circular-gran-def}}

\begin{definition}\label{def:circular}
A {\bf circular granularity} $C_{B, G}$ relates a linear granularity $G$ to the bottom granularity B, if
\begin{equation} \label{eq:circular-gran}
\begin{split}
C_{B, G}(z) & = z~\text{mod}~P(B, G) \quad \forall z \in \mathbb{Z}_{\geq 0} \\
\end{split}
\end{equation}
where
z denotes the index set,
B denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G$ with regular mapping, and $P \equiv P(B, G)$ is the number of granules of $B$ in each granule of $G$, also called the period of the grouping $(B, G)$.
\end{definition}

\begin{example}
{\rm \autoref{fig:circular-dow} illustrates the linear and the corresponding cyclical granularities. Cyclical granularities can be considered to be cutting the linear granularity into pieces, and stacking them to match the cycles (as shown in b). $B, G, H$ (day, week, fortnight, respectively) are linear granularities. The circular granularity $C_{B , G}$ (day-of-week)is constructed from $B$ and  $G$. The circular granularity $C_{B, H}$ (day-of-fortnight) is constructed from $B$ and $H$. These overlapping cyclical granularities share elements from the linear granularity. Each of $C_{B , G}$ and $C_{B , H}$ consist of repeated patterns  $\{0, 1, 2, \dots, 6\}$ and $\{0, 1, 2, \dots, 13\}$ with $P=7$ and $P=14$ respectively. Each circular granularity can use descriptive label mappings. Suppose ${L}$ is a label mapping that defines an unique label for each index $l \in \{ 0,1,\dots, (P-1)\}$, then the label mapping $L$ for $C_{B, G}$ can be defined as
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sun}, \text{Mon}, \dots, \text{Sat}\}
$$
or
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}
$$
for example.}
\end{example}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Selecting and ranking interesting pairs of cyclic temporal granularities},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

\title{Selecting and ranking interesting pairs of cyclic temporal granularities}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Exploratory data analysis, as coined by John W. Tukey (Tukey 1965) involves many iterations of finding structures and patterns that allows the data to be informative. With temporal data available at finer scales, exploring periodicity and their relationships can become overwhelming with so many possible cyclic temporal granularities (Gupta et al. 2020) to explore.

@@

{\centering \includegraphics[width=\textwidth]{figure/calendar-elec-1}

@@

\caption{Calendar display.}\label{fig:calendar-elec}
\end{figure}

\noindent Take the example of the calendar display of electricity smart meter data (\ref{fig:calendar-elec}) used in Wang, Cook, and Hyndman (2020) for four households in Melbourne, Australia. The authors show how hour-of-the-day interact with weekday and weekends and then move on to use calendar display to show daily schedules. The calendar display has several components in it, which helps us look at energy consumption across hour-of-the-day, day-of-the-week, week-of-the-month, and month-of-the-year at once. Some interaction of these cyclic granularities could also be interpreted from this display. This is a great start to have an overview of the energy consumption. However, if one wants to understand the periodicities in energy behavior and how the periodicities interact in greater details, it is not easy to comprehend the interactions of some periodicities' from this display, due to the combination of linear and cyclic representation of time. For example, this display might not be the best to understand how hour-of-the-day varies and month-of-year varies across week-of-the-month. Further, it is not clear what all interactions of cyclic granularities should be read from this display as there could be many combinations that one can look at. Moreover, calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions (Gupta et al. (2020)) and could include other cyclic granularities like hour-of-week or day-of-fortnight, which could potentially become useful depending on the context.

{\centering \includegraphics[width=\textwidth]{figure/id2-1}

\caption{something}\label{fig:id2}
\end{figure}

{\centering \includegraphics[width=\textwidth]{figure/id4-1}

\caption{something2}\label{fig:id4}
\end{figure}

Moreover, there might be specific interactions that are interesting and others that are not and that too will vary with different households. For example, area distribution quantiles are plotted for household 2 and 4 in Figure \ref{fig:id2}a and b respectively. For the first household, the 75th and 90th percentile for Jan, Feb and July are very close, implying that energy usage for these months are generally on a much higher side due to the usage of air conditioners (in Jan and Feb) and heaters (in July). The energy consumption for household 2 is also higher relative to its own consumption for Jan, Feb and March but the 75th and 90th percentile are apart implying that contrary to the first household, the second household resorts to air conditioners and heaters much less regularly than the first one. Moreover, the 75th percentile distribution is not bimodal across hours of the day for the first household in those months, but the distribution looks similar for all months for the second household. Difference in the energy consumption seem to be varying both across month-of-year (facets) and hour-of-day (x-axis). And thus, both the cyclic granularities would deem important while studying the periodicities in the first household. However, it seems like energy consumption across hours of the day are not that different across different months for the second household. Differences seem to be more prominent across month-of-year (facets) than hour-of-day (x-axis). Again, look at \ref{fig:} c and d, where energy consumption for these two households are plotted against (weekend/weekday, week-of-month). Here, for both households, the pattern of energy consumption vary across different weeks of the month irrespective of the fact it is a weekday or weekend. In that respect, the harmony pair (month-of-year, hour-of-day) seems to be more informative than (weekend/weekday, week-of-month) for the first household. It could be immensely useful to make the transition from all possible ways to only ways that could potentially be informative given a household.

The paper Gupta et al. (2020) describes how we can compute all possible combinations of cyclic time granularities. If we have \(n\) periodic linear granularities in the hierarchy table, then \(n(n-1)/2\) circular or quasi-circular cyclic granularities could be constructed. Let \(N_C\) be the total number of contextual circular, quasi-circular and aperiodic cyclic granularities that can originate from the underlying periodic and aperiodic linear granularities. The mapping of the graphical elements chosen in the paper implies that, for a numeric response variable, the graphics display distributions across combinations of cyclic granularities, one placed at x-axis and the other on the facet. That essentially implies there are \(^{N_C}P_2\) possible pairwise plots exhaustively, where each plot would display a pair of cyclic granularities. This is large and overwhelming for human consumption.

This is similar to Scagnostics (Scatterplot Diagnostics) by Tukey and Tukey (1988), which is used to discern meaningful patterns in large collections of scatterplots. Given a set of \(v\) variables, there are \(v(v-1)/2\) pairs of variables, and thus the same number of possible pairwise scatterplots. Therefore
for even small \(v\), the number of scatterplots can be
large, and scatterplot matrices (SPLOMs) could easily run out of pixels when presenting high-dimensional data. Dang and Wilkinson (2014) and Wilkinson, Anand, and Grossman (2005) provides potential solutions to this, where few characterizations help us to locate anomalies for defining several measures aimed to detect anomalies in density, shape, trend, and
other features in the 2D point scatters.

The paper (Gupta et al. (2020)) narrows down the search from \(^{N_C}P_2\) plots by identifying pairs of granularities that can be meaningfully examined together (a ``harmony''), or when they cannot (a ``clash''). However, even after excluding clashes, the list of harmonies left could be enormous for exhaustive exploration. Hence, there is a need to reduce the search even further by including only those harmonies which are informative enough. Also, ranking the remaining harmony pairs based on how well they capture the variation in the measured variable could be potentially useful.

In this paper, we aim to build a new measure to follow through these two main objectives:

\begin{itemize}
\tightlist
\item
  To choose harmonies for which distributions of categories are significantly different
\item
  To rank the selected harmonies from highest to lowest variation in the distribution of their categories.
\end{itemize}

\hypertarget{median-maximum-pairwise-distances-mmpd}{%
\section{Median Maximum Pairwise Distances (MMPD)}\label{median-maximum-pairwise-distances-mmpd}}

We are interested in assessing structure in probability distributions of the measured variable across bivariate cyclic granularities. We propose a measure called Median Maximum Pairwise Distances (MMPD) to evaluate structure in such a design.

\hypertarget{idea}{%
\subsection{Idea}\label{idea}}

The principle employed for building a new metric is explained through a simple example explained in Figure \ref{fig:null4by2}. Each of these figures have the same panel design with 2 x-axis categories and 4 facet levels. Figure \ref{fig:null4by2}a has all x categories drawn from N(5, 10) distribution for each facet. It is not an interesting display particularly, as distributions do not vary across x-axis or facet categories. Figure \ref{fig:null4by2}b has x categories drawn from the same distribution within a facet and different for different facet categories. Figure \ref{fig:null4by2}b exhibits an exact opposite situation where distribution between the x-axis categories within each facet is different but they are same across facets. Figure \ref{fig:null4by2}d takes a step further by varying the distribution across both facet and x-axis categories. If we are asked to rank the displays in order of importance from minimum to maximum, we might order it as a, b, c and then d.~It might be argued that it is not clear if b should precede or succeed c.~Gestalt theory suggests that when items are placed in close proximity, people assume that they are in the same group because they are close to one another and apart from other groups. Hence, displays that capture more variation within different categories in the same group would be important to bring out different patterns of the data. With this principle, display b could be considered less informative as compared to display c.

With reference to the graphical design in \ref{@Gupta2020-vo}, therefore the idea would be to rate a harmony pair higher if the variation between different levels of the x-axis variable is higher on an average across all levels of the facet variables. Thus the metric could be obtained by computing maximum pairwise distances between distributions of the continuous random variable across x-axis categories for all facets and then taking the median of those maximum pairwise distances across facets. This would help capture the average maximum difference in distribution of the measurement variable explained by the two cyclic granularities together. We call this metric MMPD which stands for Median Maximum Pairwise Distances. In the next section we shall see how we go about computing this measure.

{\centering \includegraphics[width=\textwidth]{figure/null4by2-1}

\caption{ A graphical display with two categories mapped to x-axis and 4 categories mapped to facets with the distribution of a continuous random variable plotted on the y-axis. Display a is not interesting as the distribution of the continuous rv does not depend across x-axis or facet categories. Display b and c are more interesting than a since there is a change in distribution either across facets(b) or x-axis(a). Display d is most interesting as distribution of the rv changes across both facet and x-axis variable.}\label{fig:null4by2}
\end{figure}

@@

\hypertarget{notations}{%
\subsubsection{Notations}\label{notations}}

Consider two cyclic granularities \(C_i\) and \(C_j\), such that \(C_i\) maps index set to a set \(\{A_k \mid k=1,\dots,K\}\) and \(C_j\) maps index set to a set \(\{B_\ell \mid \ell =1,\dots,L\}\). Here, \(A_k\) and \(B_\ell\) are the levels/categories corresponding to \(C_i\) and \(C_j\) respectively. Let \(S_{k\ell}\) be a subset of the index set such that for all \(s \in S_{k\ell}\), \(C_i(s) = A_k\) and \(C_j(s) = B_\ell\). There are \(KL\) such data subsets, one for each combination of levels (\(A_k\), \(B_\ell\)). Moreover, consider that in the graphical space, \(C_i\) is mapped to facets and \(C_j\) is mapped to x-axis.

\hypertarget{characterising-distribution}{%
\subsubsection{Characterising distribution}\label{characterising-distribution}}

Each of the \(KL\) data subsets in the data structure have multiple observations and may vary widely across different subsets due to the structure of the calendar, missing observations or uneven locations of events in the time domain. The set of observations corresponding to each combination is assumed to be a sample from an unknown probability density function.
While the whole population of observations has certain characteristics, we can typically never measure all of them. Often shape, central tendency, and variability are the common characteristics used to describe the distribution. Another way to describe the probability distribution is through quantiles. (Define quantiles here) Sample quantiles could be thought to estimate the population quantiles. But there are a large number of different definitions used for sample quantiles. The median-unbiased estimator is recommended (Rob's paper) because of its desirable properties of a quantile estimator and can be defined independently of the underlying distribution.

\hypertarget{distance-between-distributions}{%
\subsubsection{Distance between distributions}\label{distance-between-distributions}}

The most common divergence measure between distributions is the Kullback-Leibler (KL) divergence(Kullback and Leibler 1951) introduced by Solomon Kullback and Richard Leibler in 1951. The KL divergence, denoted \(D(p(x), q(x))\) is a non-symmetric measure of the difference between two probability distributions \(p(x)\) and \(q(x)\) and is interpreted as the amount of information lost when \(q(x)\) is used to approximate \(p(x)\). Although the KL divergence measures the ``distance'' between two distributions, it is not a distance measure since it is not symmetric and does not satisfy the triangle inequality. The Jensen-Shannon divergence (Menéndez et al. 1997) based on the Kullback-Leibler divergence is symmetric and it always has a finite value. The square root of the Jensen-Shannon divergence is a metric, often referred to as Jensen-Shannon distance. Other common measures of distance are Hellinger distance, total variation distance and Fisher information metric.

In the context of this paper, the pairwise distances between the distributions of the measured variable are computed through Jensen-Shannon distance (JSD) which is based on Kullback-Leibler divergence and is defined by,

\[JSD(P||Q) = \frac{1}{2}D(P||M) + \frac{1}{2}D(Q||M)\]
where \(M = \frac{P+Q}{2}\) and
\(D(P||Q) := \int^\infty_{-\infty} p(x)f(\frac{p(x)}{q(x)})\) is the KL divergence between distributions \(p(x)\) and \(q(x)\). Probability distributions are estimated through quantiles instead of kernel density so that there is minimal dependency on selecting kernel or bandwidth.

\hypertarget{algorithm}{%
\subsubsection{Algorithm}\label{algorithm}}

The algorithm employed for computing the distance measure is summarized as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Fix harmony pair \((C_i, C_j)\).
\item
  Fix \(k\). Then there are \(L\) groups corresponding to level \(A_k\) of \(C_i\).
\item
  Compute \(m = \binom{L}{2}\) pairwise distances between distributions of \(L\) unordered levels and \(m = L-1\) pairwise distances for \(L\) ordered categories.
\item
  Identify maximum within the \(m\) computed distances.
\item
  Compute maximum distance (\(M\)) using appropriate norming constants.
\item
  Use Steps 1-5 to compute maximum distance for \(\forall k \in \{1, 2, \ldots, K\}\).
\item
  Compute the distance measure MMPD\_raw = median \((M_1, M_2, \dots, M_K)\).
\end{enumerate}

A stronger measure ``max'' is chosen for aggregating x-axis categories compared to ``median'' for aggregating facet categories as the measure is intended to put more importance in pointing towards distributional differences between x-axis categories than for facet categories.

\hypertarget{distribution}{%
\subsubsection{Distribution}\label{distribution}}

From Figure \ref{fig:dist-across-facets} and \ref{fig:dist-across-x} shows the MMPD distribution is different for different levels of facets and x-axis levels. With increasing number of facets, the location of the distribution shift rightwards and with increasing x-axis levels, the scale of the distribution reduces.

Median and maximum distances are affected by the number of categories considered and hence the distance measure \(MMPD_{raw}\), which is a combination of median and maximum would also be influenced by the number of levels. It would have higher values if \(C_i\) or \(C_j\) has higher levels. We would ideally want a higher value of the measure if there is significant difference between distributions across facet or x-axis categories, and not because the number of categories are higher. In Figures \ref{fig:dist-across-facets} and \ref{fig:dist-across-x}, both the cyclic granularities \(C_i\) and \(C_j\) are considered such that their levels vary in the range \((2, 5,7 ,9, 14, 21, 30, 45)\). Each of these combinations is considered a panel for each of which \(MMPD_{raw}\) has been constructed \(100\) times to compute the distribution.
